---
title: "tidyverse_un_clj"
output: html_notebook
---

1.  Using the `read_csv()` function, read the GDP dataset into your notebook as a tibble called `gdp_df`. After reading it in, inspect the first 10 rows and then inspect the last 10 rows.

```{r}
library(tidyverse)
library(dplyr)
library(corrr)
library(extrafont)
```

Loads in tidyverse package and dplyr package

```{r}
gdp_df <- read_csv('data/gdp_per_capita.csv')

gdp_df |> 
  head(10)

gdp_df |> 
  tail(10)
```

Reads in dataframe as tibble, displays the first and last 10 rows

2.  Drop the 'Value Footnotes' column, and rename the remaining columns to 'Country', 'Year', and 'GDP_Per_Capita'.

```{r}
gdp_df <- gdp_df |> 
  select(-`Value Footnotes`)
```

Removes 'value footnotes' column

```{r}
gdp_df <- gdp_df |> 
  rename(Country=`Country or Area`,`GDP Per Capita`=Value) |> 
  mutate(Year = as.numeric(Year))
```

Renames the 1st and 3rd columns to 'Country' and 'GDP Per Capita', respectively

```{r}
year_count <- gdp_df |> 
  count(Year)
print(year_count)
```

Counts the frequency of occurrences of observations in each year and returns a tibble with that information.

Creates a scatter plot visualizing the tibble created above.

```{r}
ggplot(year_count, aes(x = Year, y = n)) +
  geom_point() +
  xlab('Year') +
  ylab('Observations') +
  labs(title = 'Amount of observations per year')
```

3.  Which years are represented in this dataset? How many observations are there per year? Make a plot to view the number of observations per year.

The years 1990 to 2021 are represented in this dataset.

```{r}
gdp_df |> 
  select(Year) |> 
  summarize(`first year` = last(Year),
            `last year` = first(Year))
```

4.  How many countries are represented in this dataset? Which countries have the fewest observations?

    242 countries, Countries listed have the least observations

```{r}
length(unique(gdp_df$Country))

gdp_df |> 
  count(Country) |> 
  arrange((n)) |> 
  head(10)
```

5.  Create a new tibble by subsetting `gdp_df` to just the year 2021. Call this new tibble `gdp_2021`

6.  Use `summary()` to find the summary statistics for GDP per capita in 2021..

```{r}
gdp_2021 <- gdp_df |> 
  filter(Year == 2021)

summary(gdp_2021)
```

7.  Create a histogram of GDP Per Capita numbers for 2021 (you may wish to adjust the number of bins for your histogram).

```{r}
ggplot(gdp_2021, aes(x=`GDP Per Capita`)) +
  geom_histogram(bins = 20, fill = "#86525B") +
  labs(title = 'Frequency of GDP Per Capita')
```

8.  Find the top 5 counties and bottom 5 countries by GDP per capita in 2021.

```{r}
gdp_2021 |> 
  arrange(desc(`GDP Per Capita`)) |>
  head(5)

gdp_2021 |> 
  arrange(`GDP Per Capita`) |>
  head(5)
```

9.  Now, return to the full dataset, `gdp_df`. Pivot the data for 1990 and 2021 (using `pivot_wider()` function) so that each row corresponds to a country, each column corresponds to a year, and the values in the table give the GDP_Per_Capita amount. Drop any rows that are missing values for either 1990 or 2021. Save the result to a tibble named `gdp_pivoted`.

```{r}
gdp_pivoted <- gdp_df |> 
  pivot_wider(names_from=Year,
              values_from =`GDP Per Capita`)

gdp_pivoted <- gdp_pivoted[, -c(3:33)]
gdp_pivoted <- gdp_pivoted |> 
  drop_na()

gdp_pivoted <- gdp_pivoted |> 
  mutate(percent_change = 100 * (`2021` - `1990`)/`1990`)
```

10. Create a new column in `gdp_pivoted` named `Percent_Change`. This column should contain the percent change in GDP_Per_Capita from 1990 to 2021. Hint: Percent change is calculated as 100\*(New Value - Old Value) / Old Value.

```{r}
nrow(gdp_pivoted |> 
  filter(percent_change < 0))

head(gdp_pivoted |> 
  arrange(desc(percent_change)), 2)
```

11. How many countries experienced a negative percent change in GDP per capita from 1990 to 2021?

    19 Countries

12. Which two countries had the highest % change in GDP per capita? Create a line plot showing these country's GDP per capita for all years for which you have data. Put both line charts on the same plot.

    Equatorial Guinea has the highest % change.

```{r}
gdp_hi <- gdp_df |> 
  filter(Country=='Equatorial Guinea')
gdp_hi2 <- gdp_df |> 
  filter(Country=='China')
```

```{r}
ggplot() +
  geom_line(data=gdp_hi,
            aes(x=Year,
                y=`GDP Per Capita`, 
                color='Equatorial Guinea')) +
  geom_line(data=gdp_hi2,
            aes(x=Year,
                y=`GDP Per Capita`,
                color='China')) +
  labs(title = 'GDP Per Capita by Year for the two highest in\n percent change from 1990 to 2021')
```

13. Read in continents.csv contained in the `data` folder into a new tibble called `continents`. We will be using this tibble to add a new column to our dataset.

```{r}
continents <- read_csv('data/continents.csv')
```

14. Merge gdp_df and continents. Keep only the countries that appear in both data frames. Save the result back to gdp_df.

```{r}
gdp_df <- gdp_df |> 
  inner_join(continents) |> 
  relocate(Continent, .before = Country) |> 
  mutate(Year = as.numeric(Year))
```

15. Determine the number of countries per continent. Create a bar chart showing this.

```{r}
cont_count <- gdp_df |> 
  group_by(Continent) |> 
  summarize(count = n())
print(cont_count)
```

```{r}
ggplot(cont_count, aes(x=Continent, y=`count`, fill=Continent)) +
  geom_col() +
  scale_fill_manual(values = c("#6C2CDB",
                               "#3D59DB",
                               "#86525B",
                               "#5C5740",
                               "#4A405C",
                               "#40485C")) +
  xlab('Continents') +
  ylab('Countries per Continent') +
  labs(title = 'Amount of Countries in each Continent')
```

16. Create a boxplot showing GDP per capita in 2021 split out by continent. What do you notice?

    Asia has the widest interquartile range, with a skewed median while Europe has the highest outliers. South America has no outliers which I find interesting.

    ```{r}
    gdp_2021 <- gdp_df |> 
      filter(Year == 2021) |> 
      group_by(Continent)
    ```

```{r}
ggplot(gdp_2021, aes(x = Continent, y = `GDP Per Capita`, fill=Continent)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#6C2CDB",
                               "#3D59DB",
                               "#86525B",
                               "#5C5740",
                               "#4A405C",
                               "#40485C")) +
  labs(title = 'GDP Per Capita of each Continent')
```

17. Read life_expectancy.csv into a tibble named life_expectancy. Do not modify the csv file in order to read this data in.

```{r}
life_expectancy <- read_csv('data/life_expectancy.csv', skip=4)
```

18. Drop the Country Code, Indicator Name, and Indicator Code columns. Then use `pivot_longer` to convert your data from wide to long. That is, instead of having one row per country and multiple colums per year, we want to have multiple rows per country and a single column for year. After melting, rename the columns to `Country`, `Year`, and `Life_Expectancy`.

```{r}
life_expectancy <- life_expectancy |> 
  select(-c(`Country Code`,
            `Indicator Name`,
            `Indicator Code`,
            `2023`,
            `...69`))
```

```{r}
life_expectancy <- life_expectancy |> 
  pivot_longer(!`Country Name`)
```

```{r}
life_expectancy <- life_expectancy |> 
  rename(Country=`Country Name`,
         Year=name,
         Life_Expectancy=value) |> 
  mutate(Year = as.numeric(Year))
```

19. What was the first country with a life expectancy to exceed 80?

    Japan in 1996 was the first year, hitting 80.2

```{r}
life_expectancy |> 
  group_by(Year) |> 
  filter(Life_Expectancy >= 80) |> 
  slice(1) |>
  head(1)
```

20. Merge `gdp_df` and `life_expectancy`, keeping all countries and years that appear in both tibbles. Save the result to a new tibble named `gdp_le`. If you get any errors in doing this, read them carefully and correct them. Look at the first five rows of your new data frame to confirm it merged correctly. Also, check the last five rows to make sure the data is clean and as expected.

```{r}
gdp_le <- gdp_df |> 
  right_join(life_expectancy) 
```

```{r}
gdp_le |> 
  head(5)

gdp_le |> 
  tail(5)
```

21. Create a new tibble, named `gdp_le_2021` by extracting data for the year 2021 from `gdp_le`. How many countries have a life expectancy of at least 80 in 2021?

There are 40 countries whose life expectancy exceeds 80 in 2021

```{r}
gdp_le_2021 <- gdp_le |> 
  filter(Year == 2021)
```

```{r}
gdp_le_2021 |> 
  filter(Life_Expectancy >= 80) |> 
  count()
```

22. Find the countries that had the top 3 largest GDP per capita figures for 2021. Create a plot showing the change in life expectancy over time for these three countries. This plot should be faceted so that each country is contained in its own figure.

```{r}
gdp_le_2021 |> 
  arrange(desc(`GDP Per Capita`)) |>  
  head(3)
```

```{r}
gdp_top3 <- gdp_le |> 
  filter(Country == c('Luxembourg', 'Singapore', 'Ireland')) |> 
  mutate(Year = as.numeric(Year))
```

```{r}
ggplot(gdp_top3, aes(x=Year, y=Life_Expectancy, group=1)) +
  geom_line() +
  facet_grid(~ Country) +
  ylab('Life Expectancy') +
  labs(title = 'Top 3 countries\' Life Expectancies\n from 1960 to 2020')
```

23. Create a scatter plot of Life Expectancy vs GDP per Capita for the year 2021. What do you notice?

As the demographic ages past 75, the gdp per capita starts to rise. Countries with longer life expectancies have more wealth.

```{r}
ggplot(gdp_le_2021, aes(x=Life_Expectancy, y=`GDP Per Capita`)) +
  geom_point(color='#40485C') +
  xlab('Life Expectancy') +
  labs(title = 'Change in GDP Per Capita by Life Expectancy') 
```

24. Find the correlation between Life Expectancy and GDP per Capita for the year 2021. What is the meaning of this number?

This number signifies that there is a strong positive relationship between an increase of gdp and an increase in life expectancy. Whether one causes the other or a third factor influences both is unknown.

```{r}
gdp_le_2021 |> 
  select(`GDP Per Capita`, Life_Expectancy) |> 
  drop_na(c('GDP Per Capita', Life_Expectancy)) |> 
  cor()
```

25. Add a column to `gdp_le_2021` and calculate the logarithm of GDP per capita. Find the correlation between the log of GDP per capita and life expectancy. How does this compare to the calculation in the previous part? Look at a scatter plot to see if the result of this calculation makes sense.

Because this is a log scale, the correlation is stronger as the data are now weighed in an evenly distributed way. Since it is a log scale, it appears that their relationship is stronger.

```{r}
gdp_le_2021 <- gdp_le_2021 |> 
  mutate(log_gdp = log(`GDP Per Capita`)) |> 
  relocate(log_gdp, .after = `GDP Per Capita`)
```

```{r}
gdp_le_2021 |> 
  select(log_gdp, Life_Expectancy) |>
  correlate()
```

```{r}
ggplot(gdp_le_2021, aes(x=Life_Expectancy, y=log_gdp)) +
  geom_point(color='#40485C') +
  geom_smooth(method = 'lm', color='#86525B') +
  xlab('Life Expectancy') +
  ylab('logarithm of GDP Per Capita') +
  labs(title = 'Change in the log of GDP Per Capita by Life Expectancy') +
  theme(text = element_text(family = "Merriweather-Italic"))
```
